---
title: "Installation and Configuration Guide"
author: "Rodolfo Tasso Suazo"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Installation and Configuration Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Installation

The simplest way to install `ciecl` is with the [`pak`](https://pak.r-lib.org/) package, which automatically resolves R and system dependencies.

### From CRAN (stable version)

```{r eval=FALSE}
install.packages("ciecl")
```

### From GitHub (development version)

```{r eval=FALSE}
install.packages("pak")
pak::pak("RodoTasso/ciecl")
```

To also install all optional dependencies (comorbidities, GT tables, ICD-11 API):

```{r eval=FALSE}
pak::pak("RodoTasso/ciecl", dependencies = TRUE)
```

## ICD-11 API Configuration (optional)

The CIE-10 functions (the core of the package) work without credentials. You only need free WHO credentials to use `cie11_search()`, which queries the WHO ICD-11 classification.

> **Security:** never write the literal key in your scripts (not even via
> `api_key = "..."`): sharing the code would expose your credentials.
> The recommended way is the `ICD_API_KEY` environment variable, set with
> `usethis::edit_r_environ()` (Option B) or via `keyring` (Option A). The
> `api_key` argument of `cie11_search()` exists only for exceptional cases
> (e.g. multiple keys or environments where env vars cannot be set).

### Step 1: Get credentials

1. Visit <https://icd.who.int/icdapi>
2. Register with your email (free)
3. You will receive a `Client ID` and a `Client Secret`

### Step 2: Store the credentials

**Option A: `keyring` (recommended)**

The [`keyring`](https://keyring.r-lib.org/) package stores secrets in the OS native keychain (macOS Keychain, Windows Credential Store, Linux Secret Service), avoiding plain-text secrets in `.Renviron`.

```{r eval=FALSE}
# Once: store "client_id:client_secret" in the keychain
keyring::key_set("ciecl_icd11")

# In each session where you use the API
Sys.setenv(ICD_API_KEY = keyring::key_get("ciecl_icd11"))
```

**Option B: `.Renviron` file**

Create or edit `~/.Renviron` (for example with `usethis::edit_r_environ()` if you have `usethis` installed) and add:

```
ICD_API_KEY=your_client_id:your_client_secret
```

Restart R for it to take effect. Do not commit `.Renviron` to Git.

**Option C: Current session only (temporary)**

```{r eval=FALSE}
Sys.setenv(ICD_API_KEY = "your_client_id:your_client_secret")
```

## SQLite cache

The package uses a local SQLite database for efficient searches. It is created automatically on first use. To force a rebuild (for example, after updating the package):

```{r eval=FALSE}
library(ciecl)
cie10_clear_cache()
```

## Verify the installation

```{r eval=FALSE}
library(ciecl)

# Check that the package loads correctly
packageVersion("ciecl")

# Verify catalogue access
nrow(cie10_cl)

# Test a basic lookup
cie_lookup("E11.0")

# Test fuzzy search
cie_search("diabetes")
```

## Support

- Report issues: <https://github.com/RodoTasso/ciecl/issues>
- Documentation: <https://rodotasso.github.io/ciecl/>
